---
title: "Diabetes Death Dashboard"
author: "Ruigang Jiang"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(plotly)
library(scales)
library(readr)
library(flexdashboard)
library(DT)
```

```{r data-prep, include=FALSE}
diabetes_data <- read_csv("AH_Provisional_Diabetes_Death_Counts_for_2020.csv")

diabetes_clean <- diabetes_data %>%
  mutate(
    Date_Of_Death_Year = as.numeric(Date_Of_Death_Year),
    Date_Of_Death_Month = as.numeric(Date_Of_Death_Month),
    Diabetes.uc = as.numeric(Diabetes.uc),
    Diabetes.mc = as.numeric(Diabetes.mc),
    C19PlusDiabetes = as.numeric(C19PlusDiabetes),
    month_label = month.abb[Date_Of_Death_Month]
  ) %>%
  filter(
    !is.na(Date_Of_Death_Month),
    !is.na(AgeGroup),
    !is.na(Sex)
  )

monthly_deaths <- diabetes_clean %>%
  group_by(Date_Of_Death_Month, month_label) %>%
  summarise(
    total_diabetes_uc = sum(Diabetes.uc, na.rm = TRUE),
    total_diabetes_mc = sum(Diabetes.mc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Date_Of_Death_Month)

age_sex_deaths <- diabetes_clean %>%
  group_by(AgeGroup, Sex) %>%
  summarise(
    total_diabetes_uc = sum(Diabetes.uc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    !AgeGroup %in% c("All ages", "Not stated"),
    !Sex %in% c("All Sexes")
  ) %>%
  mutate(
    AgeGroup = factor(
      AgeGroup,
      levels = c(
        "<18 years", "18-29 years", "30-39 years", "40-49 years",
        "50-59 years", "60-69 years", "70-79 years", "80 years and over"
      )
    )
  ) %>%
  arrange(AgeGroup, Sex)
```

Overview
=====================================

Row
-------------------------------------

### Dashboard Summary

This dashboard explores provisional diabetes-related death counts in the United States in 2020. The visualizations highlight monthly mortality trends and demographic differences across age groups and sex.

### Why This Matters

Understanding when diabetes-related deaths increased and which groups were most affected can support public health surveillance and targeted intervention planning.

Row
-------------------------------------

### Monthly Diabetes Death Trends in 2020

*Deaths peak in April and remain elevated mid-year; multiple-cause counts are consistently higher than underlying-cause counts.*

```{r monthly-plot}
plot_ly(
  data = monthly_deaths,
  x = ~Date_Of_Death_Month,
  y = ~total_diabetes_uc,
  type = "scatter",
  mode = "lines+markers",
  name = "Underlying Cause",
  text = ~paste0(
    "Month: ", month_label,
    "<br>Underlying deaths: ", comma(round(total_diabetes_uc, 0))
  ),
  hoverinfo = "text",
  line = list(width = 3),
  marker = list(size = 8)
) %>%
  add_trace(
    y = ~total_diabetes_mc,
    mode = "lines+markers",
    name = "Multiple Cause",
    text = ~paste0(
      "Month: ", month_label,
      "<br>Multiple deaths: ", comma(round(total_diabetes_mc, 0))
    ),
    hoverinfo = "text",
    line = list(width = 3),
    marker = list(size = 8)
  ) %>%
  layout(
    title = list(
      text = "Monthly Diabetes Death Counts",
      y = 0.95
    ),
    xaxis = list(
      title = "Month of Death in 2020",
      tickmode = "array",
      tickvals = 1:12,
      ticktext = month.abb
    ),
    yaxis = list(title = "Number of Deaths"),
    legend = list(
      orientation = "h",
      x = 0.3,
      y = 1.05
    ),
    margin = list(
      l = 70,
      r = 30,
      t = 80,
      b = 70
    )
  )
```

Row
-------------------------------------

### Diabetes Deaths by Age Group and Sex

*Mortality rises steeply with age and is consistently higher among males across most age groups.*

```{r age-sex-plot}
plot_ly(
  data = age_sex_deaths,
  x = ~AgeGroup,
  y = ~total_diabetes_uc,
  color = ~Sex,
  type = "bar",
  text = ~paste0(
    "Age group: ", AgeGroup,
    "<br>Sex: ", Sex,
    "<br>Deaths: ", comma(round(total_diabetes_uc, 0))
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = list(
      text = "Underlying-Cause Diabetes Deaths by Age Group and Sex",
      y = 0.95
    ),
    xaxis = list(
      title = "Age Group",
      tickangle = -45
    ),
    yaxis = list(title = "Number of Deaths"),
    barmode = "group",
    legend = list(
      orientation = "h",
      x = 0.3,
      y = 1.05,
      title = list(text = "Sex")
    ),
    margin = list(
      l = 70,
      r = 30,
      t = 80,
      b = 100
    )
  )
```

Dataset
=====================================

Row
-------------------------------------

### Dataset Description

This dashboard uses the CDC provisional diabetes death counts dataset for 2020.
The dataset includes deaths in which diabetes was recorded as an underlying cause or as a multiple cause of death.
These records were collected through national mortality reporting systems in the United States.
The study population is the U.S. population represented in provisional death reporting.
The dashboard focuses on deaths reported during calendar year 2020.

Row
-------------------------------------

### Notes for Interpretation

Because these are provisional counts, they may differ from finalized mortality statistics released later.
The charts in this dashboard are intended for descriptive exploration of temporal and demographic patterns rather than causal inference.

Row
-------------------------------------

### Sample of Raw Data
The table below shows the first 50 rows of the cleaned CDC provisional diabetes death counts dataset used in this dashboard.

```{r data-display}
DT::datatable(
  diabetes_clean %>% head(50),
  options = list(pageLength = 10, scrollX = TRUE)
)
```
Repository
=====================================

Row
-------------------------------------

### GitHub Repository

Source code for this dashboard is available here:
[GitHub Repository](https://github.com/mihotelk/data555-final-dashboard)

Row
-------------------------------------

### Project Relevance

Understanding when diabetes-related mortality increased and which demographic groups were most affected can help guide surveillance and intervention priorities.
A simple interactive dashboard makes these patterns easier to communicate to both technical and non-technical audiences.


























