---
title: "Diabetes Death Dashboard"
author: "Ruigang Jiang"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(plotly)
library(scales)
library(readr)
library(flexdashboard)
```

```{r data-prep, include=FALSE}
diabetes_data <- read_csv("AH_Provisional_Diabetes_Death_Counts_for_2020.csv")

diabetes_clean <- diabetes_data %>%
  mutate(
    Date_Of_Death_Year = as.numeric(Date_Of_Death_Year),
    Date_Of_Death_Month = as.numeric(Date_Of_Death_Month),
    Diabetes.uc = as.numeric(Diabetes.uc),
    Diabetes.mc = as.numeric(Diabetes.mc),
    C19PlusDiabetes = as.numeric(C19PlusDiabetes),
    month_label = month.abb[Date_Of_Death_Month]
  ) %>%
  filter(
    !is.na(Date_Of_Death_Month),
    !is.na(AgeGroup),
    !is.na(Sex)
  )

monthly_deaths <- diabetes_clean %>%
  group_by(Date_Of_Death_Month, month_label) %>%
  summarise(
    total_diabetes_uc = sum(Diabetes.uc, na.rm = TRUE),
    total_diabetes_mc = sum(Diabetes.mc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Date_Of_Death_Month)

age_sex_deaths <- diabetes_clean %>%
  group_by(AgeGroup, Sex) %>%
  summarise(
    total_diabetes_uc = sum(Diabetes.uc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    !AgeGroup %in% c("All ages", "Not stated"),
    !Sex %in% c("All Sexes")
  ) %>%
  mutate(
    AgeGroup = factor(
      AgeGroup,
      levels = c(
        "<18 years", "18-29 years", "30-39 years", "40-49 years",
        "50-59 years", "60-69 years", "70-79 years", "80 years and over"
      )
    )
  ) %>%
  arrange(AgeGroup, Sex)
```

# Overview

## Row

### Diabetes Mortality in 2020

This dashboard explores provisional diabetes-related death counts in the United States during 2020.
The first chart highlights monthly trends, while the second compares deaths across age groups and sex.

### Why This Matters

This dashboard helps identify temporal and demographic patterns in diabetes-related mortality in the United States.
These findings can support public health monitoring and more targeted prevention strategies.

## Row

### Monthly Diabetes Death Trends in 2020

```{r monthly-plot}
plot_ly(
  data = monthly_deaths,
  x = ~Date_Of_Death_Month,
  y = ~total_diabetes_uc,
  type = "scatter",
  mode = "lines+markers",
  name = "Underlying Cause",
  text = ~paste0(
    "Month: ", month_label,
    "<br>Underlying-cause deaths: ", comma(round(total_diabetes_uc, 0))
  ),
  hoverinfo = "text",
  line = list(width = 3),
  marker = list(size = 8)
) %>%
  add_trace(
    y = ~total_diabetes_mc,
    mode = "lines+markers",
    name = "Multiple Cause",
    text = ~paste0(
      "Month: ", month_label,
      "<br>Multiple-cause deaths: ", comma(round(total_diabetes_mc, 0))
    ),
    hoverinfo = "text",
    line = list(width = 3),
    marker = list(size = 8)
  ) %>%
  layout(
    title = list(text = "Monthly Diabetes Death Counts"),
    xaxis = list(
      title = "Month of Death in 2020",
      tickmode = "array",
      tickvals = 1:12,
      ticktext = month.abb
    ),
    yaxis = list(title = "Number of Deaths"),
    legend = list(orientation = "h", x = 0.1, y = 1.12)
  )
```

This interactive line chart shows that deaths with diabetes listed as a multiple cause were consistently higher than deaths with diabetes listed as the underlying cause. Both measures rose sharply in spring 2020, with a visible peak around April, and then declined later in the year.

## Row

### Diabetes Deaths by Age Group and Sex

```{r age-sex-plot}
plot_ly(
  data = age_sex_deaths,
  x = ~AgeGroup,
  y = ~total_diabetes_uc,
  color = ~Sex,
  type = "bar",
  text = ~paste0(
    "Age group: ", AgeGroup,
    "<br>Sex: ", Sex,
    "<br>Underlying-cause deaths: ", comma(round(total_diabetes_uc, 0))
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = list(text = "Underlying-Cause Diabetes Deaths by Age Group and Sex"),
    xaxis = list(title = "Age Group", tickangle = -35),
    yaxis = list(title = "Number of Deaths"),
    barmode = "group",
    legend = list(title = list(text = "Sex"))
  )
```

This grouped bar chart shows that diabetes-related deaths increased strongly with age, especially in older adult groups. Interactive tooltips and legend controls make it easier to compare male and female death counts within each age group.

# Dataset

## Row

### Dataset Description

This dashboard uses the CDC provisional diabetes death counts dataset for 2020.
The dataset includes deaths in which diabetes was recorded as an underlying cause or as a multiple cause of death.
These records were collected through national mortality reporting systems in the United States.
The study population is the U.S. population represented in provisional death reporting.
The dashboard focuses on deaths reported during calendar year 2020.

## Row

### Notes for Interpretation

Because these are provisional counts, they may differ from finalized mortality statistics released later.
The charts in this dashboard are intended for descriptive exploration of temporal and demographic patterns rather than causal inference.

# Repository

## Row

## GitHub Repository

Source code for this dashboard is available here:

<a href="https://github.com/mihotelk/data555-final-dashboard" target="_blank">
https://github.com/mihotelk/data555-final-dashboard
</a>
























