---
title: "DATA 555 Final Project 3a"
author: "Ruigang Jiang"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(scales)
library(htmltools)
```

## Introduction

This document presents two interactive visualizations based on provisional diabetes-related death counts in 2020. The purpose is to explore patterns in deaths over time and compare counts across demographic groups using interactive tools.

## Data Import

```{r, message=FALSE, warning=FALSE}
diabetes_data <- read_csv("AH_Provisional_Diabetes_Death_Counts_for_2020.csv")
glimpse(diabetes_data)
```

## Data Cleaning

```{r, message=FALSE, warning=FALSE}
diabetes_clean <- diabetes_data %>%
  mutate(
    Date_Of_Death_Year = as.numeric(Date_Of_Death_Year),
    Date_Of_Death_Month = as.numeric(Date_Of_Death_Month),
    Diabetes.uc = as.numeric(Diabetes.uc),
    Diabetes.mc = as.numeric(Diabetes.mc),
    C19PlusDiabetes = as.numeric(C19PlusDiabetes),
    month_label = month.abb[Date_Of_Death_Month]
  ) %>%
  filter(!is.na(Date_Of_Death_Month),
         !is.na(AgeGroup),
         !is.na(Sex))

head(diabetes_clean)
```

## Plot 1: Interactive Line Chart of Monthly Diabetes-Related Death Trends in 2020

```{r, message=FALSE, warning=FALSE}
monthly_deaths <- diabetes_clean %>%
  group_by(Date_Of_Death_Month, month_label) %>%
  summarise(
    total_diabetes_uc = sum(Diabetes.uc, na.rm = TRUE),
    total_diabetes_mc = sum(Diabetes.mc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Date_Of_Death_Month)

p1 <- plot_ly(
  data = monthly_deaths,
  x = ~Date_Of_Death_Month,
  y = ~total_diabetes_uc,
  type = "scatter",
  mode = "lines+markers",
  text = ~paste(
    "Month:", month_label,
    "<br>Underlying diabetes deaths:", comma(total_diabetes_uc)
  ),
  hoverinfo = "text",
  name = "Underlying Cause"
) %>%
  add_trace(
    y = ~total_diabetes_mc,
    text = ~paste(
      "Month:", month_label,
      "<br>Multiple-cause diabetes deaths:", comma(total_diabetes_mc)
    ),
    hoverinfo = "text",
    name = "Multiple Cause"
  ) %>%
  layout(
    title = list(text = "Monthly Diabetes-Related Deaths in 2020", font = list(size = 18)),
    xaxis = list(
      title = "Month of Death",
      tickmode = "array",
      tickvals = monthly_deaths$Date_Of_Death_Month,
      ticktext = monthly_deaths$month_label
    ),
    yaxis = list(
      title = "Number of Deaths"
    ),
    legend = list(title = list(text = "Death Type"))
  )

p1
```

This visualization shows the monthly trend in diabetes-related deaths during 2020. Deaths where diabetes was recorded as a multiple cause are consistently much higher than deaths where diabetes was the underlying cause, and both measures peak around April before declining later in the year. The interactive hover tooltips allow users to inspect the exact number of deaths in each month and compare patterns over time.

## Plot 2: Diabetes Deaths by Age Group and Sex

```{r, message=FALSE, warning=FALSE}
age_sex_deaths <- diabetes_clean %>%
  group_by(AgeGroup, Sex) %>%
  summarise(
    total_diabetes_uc = sum(Diabetes.uc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(
    !AgeGroup %in% c("All ages", "Not stated"),
    !Sex %in% c("All Sexes")
  )

p2 <- plot_ly(
  data = age_sex_deaths,
  x = ~AgeGroup,
  y = ~total_diabetes_uc,
  color = ~Sex,
  type = "bar",
  text = ~paste(
    "Age group:", AgeGroup,
    "<br>Sex:", Sex,
    "<br>Underlying diabetes deaths:", comma(total_diabetes_uc)
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = list(
      text = "Underlying Diabetes Deaths by Age Group and Sex",
      font = list(size = 18)
    ),
    xaxis = list(
      title = "Age Group",
      tickangle = -45
    ),
    yaxis = list(
      title = "Number of Deaths"
    ),
    barmode = "group",
    legend = list(title = list(text = "Sex"))
  )

p2
```

This visualization compares underlying diabetes-related deaths across age groups for males and females in 2020. Death counts increase dramatically with age, with the largest numbers occurring among individuals aged 65 years and older, especially in the 70–79 and 80+ age groups. The interactive hover tool tips allow users to inspect the exact number of deaths for each age group and easily compare patterns between males and females.



























